---
title: "ESM 262 HW1 - Bertalanffy Function"
author: "Lauren Puffer"
format: html
editor: visual
embed-resources: true
code-fold: show
execute: 
  warning: false
  message: false
toc: TRUE
theme: journal
---

## Load packages

```{r}
library(tidyverse)
library(here)
library(janitor)
library(dplyr)
```

## Function description

The Bertalanffy Age-Length equation is used in fisheries science to determine the age of a fish using length and weight data. This function will take the weight (g) and fork length (cm) of a Coho salmon and give you the approximate age of the fish in years. The function can be applied to a data set to examine the distribution of fish in the population by their age with a bar graph.

## Create a help page with parameters

```{r}
#' Bertalanffy Age-Length Function for Coho Salmon
#' 
#' This is a function that calculates the age of a fish in years give the fork length in centimeters and the weight in grams
#' 
#' @param tl Fork length (cm)
#' @param w Weight (g)
#' @param 109.22 (cm) The maximum length of an adults Coho salmon. 
#' @param k The mean biological condition coefficient which indicates the health of the fish population. 
#' @return t (years)
```

## Bertalanffy function definition

```{r}
#create function where tl is total length (cm) and w is weight (g) in 

bertalanffy <- function(tl, w) {
  
  # Error checks for negative values
  ifelse((tl <0), return("Total length cannot be negative."), tl)
  
  ifelse((w<0), return("Weight cannot be negative."), w)
  
  #Check that both vectors are numeric
  if (!is.numeric(tl) || !is.numeric(w)) {
    return("Both total length (tl) and weight (w) must be numeric.")
  }
  
  
  k = ( 100 )*(w / (tl ^ 3)) 
  #mean biological condition factor used in Bertalanffy eqn. 
  
  t = ( - k ) * log( (109.22-tl) /109.22) 
  #avg. fry length is 2.5 cm. max adult lenth is 109.22
  
    #write error codes for uneven lengths of values
    if (length(tl) != length(w)) {
    return("Number of length values is not equal to number of weight values.")
  }
  
  return(t) #get the age of the fish
}
```

## Test function and error codes

```{r}
#test with a hypothetical fish
bertalanffy(30, 1000)

#test with negative values
bertalanffy(-30, 1000)

bertalanffy(30, -1000)

#create datasets to test length errors
fish_length <- 1:5
fish_weight <- 1:20

#test length errors
bertalanffy(fish_length, fish_weight)

```

Now that we've confirmed that the function is working and error codes are generated, we can apply the function to a data set and create an age column within the data frame with the outputs from our function.

## Data citation

**U.S. GLOBEC Northeast Pacific**

Wertheimer, A., Cokelet, E., Farley, E., Moss, J. (2009) Salmon length, weight, sex, stomach data from F/V Great Pacific, R/V Miller Freeman multiple cruises in the Coastal Gulf of Alaska, NE Pacific from 2001-2004 (NEP project). Biological and Chemical Oceanography Data Management Office (BCO-DMO). (Version 1) Version Date 2009-04-16 \[if applicable, indicate subset used\]. doi:10.1575/1912/bco-dmo.3106.1

## Import Coho salmon data set

```{r}
salmon_data <- read.csv(here("Functions HW", "data", "salmon_biodata.csv")) |>
  clean_names()
```

## Data wrangling

```{r}
#select only weight and length data for "wet" and not frozen fish
#filter for coho salmon
coho_df <- salmon_data |>
  filter(species_common_name == "Coho") |>
  select(length_wet, weight_wet)|>
  drop_na() |>
      mutate(length_wet = as.numeric(length_wet),  #convert length_wet to numeric
    weight_wet = as.numeric(weight_wet)) |>
  mutate(length_wet = (length_wet/10)) #convert lengths from mm to cm

#check for NA values coerced from 'mutate' function
non_numeric_rows <- coho_df[!grepl("^\\d*\\.?\\d*$", coho_df$weight_wet), ]
print(non_numeric_rows)

#some rows still have NAs so we must remove them 
coho_df <- coho_df |>
  filter(!is.na(as.numeric(weight_wet)))

#make sure both are numeric
is.numeric(coho_df$weight_wet)

is.numeric(coho_df$length_wet)
```

## Function with Coho salmon data set

```{r}
#specify vectors
total_length = coho_df$length_wet
weight = coho_df$weight_wet

#use function on coho salmon data set
results = bertalanffy(total_length, weight)

#create age column w results from function
coho_df$age = results
```

## Create bar graph with ages of Coho salmon

```{r}

coho_age_graph <- ggplot(coho_df, aes(x = age, y = length_wet)) +
  geom_point() +
  geom_smooth()+
  theme_classic()

print(coho_age_graph)
```

The age of Coho salmon and their length seem to have a linear relationship. This is what we would expect to see with the Bertalanffy equation.
